---
title: "intro-to-rtails"
author: "Sebastián Pardo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro-to-rtails}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)
```

## Introduction to `rtails`

The `rtails` package provides functions to easily obtain
heavy-tailed multiplicative random deviates (i.e., positive only, centered around 1).
While its primary aim is to simulate heavy-tailed recruitment deviations, these
functions could be used for other applications as well.

The heavy-tailed random deviates are generated using four distributions: 

- Normal mixture (i.e. normal with sporadic wider normal draws): `rnorm_tails()`
- Pareto: `rpareto_tails()`
- Student-t: `rst_tails()`
- Generalized Extreme Variance: `rgev_tails()`

Where possible, functions are analytically bias-corrected around a mean of 1 
by default. However, this correction is not always applicable
(e.g., Student-t and GEV). We therefore also allow all functions to 
be bias corrected using the sample mean (`sample_bias_correct`). 
All functions also take a `seed` argument so that different parameterizations 
of each distribution using a given seed are more directly comparable.

We also provide the `plot_tails()` function for easy visualization of the 
generated random deviates.
 
## Functions 

### Normal mixture: `rnorm_tails()`

The `rnorm_tails()` function gives you control over two separate normal distributions,
with the underlying non-heavy-tailed spread determined by `sigma`, and the
wider distribution determined by `high_sigma` with the frequency of 
sporadic events determined by `rate`.
The normal mixture can be used as a simple intuitive
way of parameterizing heavy-tailed recruitment, but is particularly suitable
for situations where it is thought there are two discrete states that have different
recruitment dynamics.

Aside from specifying the sample size with `n`, the `rnorm_tails()` function takes
two variance values: one for the underlying distribution (`sigma`), and the other
for the heavy-tailed sporadic values (`high_sigma`). The probability of replacing each 
value with the sporadic draw is given by the `rate` argument, while autocorrelation (`ac`), skew (`skew`),
and bias correction using the sample mean (`sample_bias_correction`) can also be incorporated.

```{r packages}
library(rtails)
library(magrittr)
```

```{r rnorm_tails}
set.seed(51)

rnt <- rnorm_tails(80, sigma = 0.1, high_sigma = 2, rate = 1/14)
plot_tails(rnt)
```

Notice that the sporadic recruitment points (i.e., those that diverge considerably
from the underlying normal distribution) are marked in red. The positions of 
these sporadic draws are stored in the `ht` attribute of the vector object:

```{r rnorm_tails_attr}
attr(rnt, "ht")
```

This function also takes as a first argument a vector of random deviates that
is updated to include sporadic events:

```{r rnorm_tails_replace}
rn <- rnorm(80, mean = -0.5 * 0.1^2, sd = 0.1) %>% exp() # bias-corrected mean
plot_tails(rn)

rnt2 <- rnorm_tails(rn, high_sigma = 2, rate = 1/10, replace = TRUE, seed = 19)
plot_tails(rnt2)
```

Note how most heavy-tailed sporadic events are biased low; this is because the 
wider normal, when exponentiated, can produce values that are extremely high, thus
to maintain a mean of one there are usually many more low points than high:

```{r rnorm_tails_replace_2}
rnt3 <- rnorm_tails(rn, high_sigma = 2, rate = 1/10, replace = TRUE, seed = 25)
plot_tails(rnt3)
```

Autocorrelation and skewness can be easily incorporated as well:

```{r rnorm_tails_ac}
rnt4 <- rnorm_tails(60, high_sigma = 2, rate = 1/10, ac = 0.7,
                    seed = 1984)
plot_tails(rnt4)
```

```{r rnorm_tails_skew}
rnt5 <- rnorm_tails(60, high_sigma = 2, rate = 1/10, skew = 20000000,
                    seed = 1984)
plot_tails(rnt5)
mean(rnt5)
```

```{r rnorm_tails_skew_2}
rnt6 <- rnorm_tails(60, high_sigma = 2, rate = 1/10, skew = -20000000,
                    seed = 1984)

plot_tails(rnt6)
mean(rnt6)
```

A negatively skewed mixture normal can be used to model heavy-tailed recruitment
failures, where the rate at which they occur can be explicitly incorporated.

### Pareto distribution: `rpareto_tails()`

The Pareto distribution function has a `shape` parameter, where
lower values result in a wider spread and heavier tails:

```{r rpareto_tails}
rpt <- rpareto_tails(80, 3)
plot_tails(rpt)

rpt2 <- rpareto_tails(80, 30)
plot_tails(rpt2)
```

Fortunately, the Pareto distribution can be bias-corrected and therefore 
the random deviates are centered around one:

```{r pareto_tails_means}
replicate(5, rpareto_tails(100000, 3) %>% mean)
```

Autocorrelation can also be included:

```{r pareto_tails-ac-skew}
rpareto_tails(80, 30, ac = 0.25, sample_bias_correct = TRUE, 
              bias_correct = FALSE) %>% plot_tails()
```

### Student-t distribution: `rst_tails()`

The `rst_tails()` function also takes one parameter for the degrees
of freedom `df`, and tends to be a predominantly right-tailed distribution
that approximates the normal as `df` increases to infinity. The 
Student-t distribution mixes frequency and severity for a general heavy-tailedness. 

```{r rst_tails}
rst <- rst_tails(80, 20, seed = 42)
plot_tails(rst)
```

```{r rst-mean}
mean(rst)
```

Notice how the sample mean is exactly one. This is because the Student-t
distribution cannot be analytically bias corrected, so a bias correction
using the sample mean is applied by default.

```{r rst_tails_nobiascorr}
rst2 <- rst_tails(80, 20, seed = 42, bias_correct = FALSE)
plot_tails(rst2)
```

Autocorrelation and skew can also be included:

```{r rst_tails-ac-skew}
rst_tails(80, 20, seed = 42, ac = 0.7) %>% plot_tails()

rst_tails(80, 20, seed = 42, skew = -1.2) %>% plot_tails()
```

Including a `skew` argument uses a different function (`fGarch::rsstd()`) so
the resulting draws will be different to those without skew even with
the same `seed`.

### Generalized extreme value distribution: `rgev_tails()`

The GEV distribution function takes two parameters: `shape` and `scale`, the 
latter which has to be greater than zero.

```{r rgev_tails}
rgev <- rgev_tails(80, scale = 0.23, shape = 0.2, seed = 32)
plot_tails(rgev)
```

Values of `scale` greater than 1 result in extremely right-tailed distributions,
perhaps much more so than any recruitment scenario:

```{r rgev_tails_2}
rgev2 <- rgev_tails(80, scale = 2, shape = 0.2)
plot_tails(rgev2)
```

Including autocorrelation:

```{r rgev_tails_ac}
rgev_tails(80, scale = 0.23, shape = 0.2, ac = 0.8) %>% plot_tails()
```

## Usage examples

Some temperate rockfishes in the genus *Sebastes* show a pattern of recruitment
where there are multiple decades of low recruitment followed by a single year
(or even a few years) of very high recruitment that can sustain the population
for quite a while as it is fished down. The yelloweye rockfish (*S. ruberrimus*) 
and Bocaccio rockfish (*S. paucispinis*) in the coast of British Columbia, Canada, 
as well as deepwater redfish (*S. mentella*) and Acadian redfish (*S. fasciatus*)
in Atlantic Canada are well-known examples. Some fast growing species also
show a similar recruitment pattern, for example Atlantic mackerel 
(*Scomber scombrus*) shows years of very high recruitment once or twice per decade.


```{r yelloweye}
rpareto_tails(60, shape = 1.5, seed = 1) %>% plot_tails()

rst_tails(60, df = 6, seed = 8) %>% plot_tails()

# sample bias correcting provides a right-skewed distribution of 
# extreme events
rnorm_tails(60, sigma = 0.1, high_sigma = 4, rate = 1/20, bias_correct = FALSE,
            sample_bias_correct = TRUE, seed = 44) %>% 
  plot_tails()
```

Pacific Cod (*Gadus macrocephalus*) in Alaska is a stock where extreme low recruitment is 
hypothesized to have been caused by a lack of spawning environment due to 
elevated bottom temperatures from 2014 to 2016. The deviations are probably not as extreme as in the above rockfish example, and the concern 
is about downward only events that sometimes appear to cycle on an approximately decadal time span.
For such a stock stock, a negatively-skewed mixture normal would be suitable
as it can incorporate both extreme low recruitment (skewness of wider distribution)
and its decadal nature (`rate = 1/10`):

```{r pacific-cod} 
rnorm_tails(60, sigma = 0.1, high_sigma = 2, rate = 1/10, 
            skew = -100, seed = 1984) %>%
  plot_tails()
```
 
Another way of thing about heavy-tail distributions is in terms of frequency of extreme
events, directionality (left- or right-skewed), and intensity (slightly 
heavy tailed vs. extremely heavy tailed). These axes of variation are controlled
by different arguments in each function:

- frequency: `rate` parameter in `rnorm_tails()`, `df` in `rst_tails()`
- direction: 
    - left (towards zero) `rnorm_tails()`with `skew`
    - right (towards infinity): `rst_tails()`,`rpareto_tails()`
- intensity: `high_sigma` parameter in `rnorm_tails()`, `df` in `rst_tails()`, `shape` parameter in `rgev_tails()` and
`rpareto_tails()`.


### Incorporating recruitment deviations to Operating Models in `SAMtool`

A primary usage of this package is to provide random recruitment deviations 
to be used for Management Strategy Evaluation (MSE) using the `MSEtool` package.
This can be done directly by providing recruitment deviates as a custom parameter,
specifically the `Perr_y` parameter:


```{r samtool-validcpars}
library(DT)
library(dplyr)
library(MSEtool)
MSEtool:::cpars_info %>% filter(Var == "Perr_y") %>% datatable()
```


The `Perr_y` custom parameter has specific dimensions that need to be followed,
however for incorporating heavy-tailed recruitment in future projections we 
have to focus on providing deviations for columns starting at `maxage + nyears + 1`.

A simple example of incorporating normal mixture random deviates using a
Student-t distribution (`rst_tails()`):


```{r example-om}
om <- new("OM", Rockfish, Generic_Fleet, Generic_Obs, Perfect_Imp)

om@nsim
om@nsim <- 5 # fewer simulations for faster runs
om@maxage
om@proyears
om@nyears

# Creating Perr_y matrix 
Perr_y <- matrix(data = 1, nrow = om@nsim, 
                 ncol = om@maxage + om@nyears + om@proyears)
dim(Perr_y)

recdevs <- rst_tails(n = om@nsim * om@proyears, df = 7.5, seed = 42) 
plot_tails(recdevs) 

Perr_y[,(om@maxage + om@nyears + 1):ncol(Perr_y)] <- recdevs

om_ht <- om # copying om
om_ht@cpars <- list(Perr_y = Perr_y) # storing recruitment deviations in cpars
```

Finally, run the MSEs:

```{r runmse, eval=FALSE}
mse <- runMSE(OM = om, MPs = c("curEref", "FMSYref"))
mse_ht <- runMSE(OM = om_ht, MPs = c("curEref", "FMSYref"))
```

**TO DO: Add two simple figures to show differences in MSEs**
