---
title: "Using the rtails package"
author: "Sebastián Pardo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)
```

## Introduction

The `rtails` package aims to provide an easy set of functions to obtain
heavy-tailed multiplicative random deviates (i.e., positive only, centered around 1), 
to expand on potential recruitment deviation scenarios.

So far, the heavy-tailed random deviates are generated using: 

- normal mixture (i.e. standard normal with sporadic wider normal draws)
- Pareto
- Student-t

Where possible, functions are bias-corrected around a mean of 1, however a 
correction is not always applicable. Both the normal mixture and Student-t
distributions are exponentiated so that values are positive and (ideally)
centered around 1.

We also provide the `plot_tails()` function, for easy visualization of the 
generated random deviates.

### Normal mixture: `rnorm_tails()`

Aside from specifying the sample size with `n`, the `rnorm_tails()` function takes
two variance values: one for the underlying distribution (`sigma`), and the other
for the heavy-tailed sporadic values (`highsigma`). The probability of replacing each 
value with the sporadic draw is given by the `rate` argument:

```{r setup}
# library(rtails)
devtools::load_all()
library(magrittr)
```

```{r rnorm_tails}
set.seed(51)

rnt <- rnorm_tails(80, sigma = 0.1, highsigma = 2, rate = 1/38)
plot_tails(rnt)
```

Notice that the sporadic recruitment points (i.e. those that diverge considerably
from the underlying normal distribution) are marked in red. The positions of 
these sporadic draws are stored in the `ht` attribute of the vector object:
```{r rnorm_tails_attr}
attr(rnt, "ht")
```

This function also takes as a first argument a vector of random deviates that
is updated to include sporadic events:

```{r rnorm_tails_replace}
rn <- rnorm(80, mean = -0.5 * 0.1^2, sd = 0.1) %>% exp() # bias-corrected mean
plot_tails(rn)

rnt2 <- rnorm_tails(rn, highsigma = 2, rate = 1/10, replace = TRUE)
plot_tails(rnt2)
```

### Pareto distribution: `rpareto_tails()`

The Pareto distribution function takes only a `shape` parameter, where
lower values result in wider tails:

```{r rpareto_tails}
rpt <- rpareto_tails(80, 3)
plot_tails(rpt)

rpt2 <- rpareto_tails(80, 30)
plot_tails(rpt2)
```

Fortunately, the Pareto distribution can be bias-corrected and therefore
the random deviates are centered around one:

```{r}
replicate(5, rpareto_tails(100000, 3) %>% mean)
```

### Student-t distribution: `rst_tails()`

The Student-t distribution function also takes one parameter for the degrees
of freedom `df`. 

```{r rst_tails}
rst <- rst_tails(80, 20)
plot_tails(rst)
```

Notice how the sample mean is considerably different from 1. However, as the 
Student-t distribution becomes the normal when `df = Inf`, we also
provide the option to bias-correct the values using the same correction factor used
for normal distributions before exponentiating (mean = $\mu-\sigma^2/2$). Obviously,
this means the bias-correction will be more accurate as the values of `df` increase:

```{r rst_tails_biascorr}
rst2 <- rst_tails(80, 20, bias.correct = TRUE)
plot_tails(rst2)
```
